Cognate beginnings to bilingual lexical acquisition

### 2023-06-28 manuscript/manuscript.tex
Add manuscript comments
Update PDF format

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The role of cross-linguistic lexical similarity on bilingual word acquisition},
  pdfauthor={Gonzalo Garcia-Castro1, Daniela Avila-Varela1, \& Nuria Sebastian-Galles1},
  pdflang={en-EN},
  pdfkeywords={lexical acquisition, vocabulary, bilingualism},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Cross-linguistic similarity and  word acquisition}
\keywords{lexical acquisition, vocabulary, bilingualism\newline\indent Word count: X}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifXeTeX
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[main=english]{babel}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\fi
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{The role of cross-linguistic lexical similarity on bilingual word acquisition}
\author{Gonzalo Garcia-Castro\textsuperscript{1}, Daniela Avila-Varela\textsuperscript{1}, \& Nuria Sebastian-Galles\textsuperscript{1}}
\date{}

\authornote{

Campus de Ciutadella, Universitat Pompeu Fabra, 08005, Barcelona, Spain

Correspondence concerning this article should be addressed to Gonzalo Garcia-Castro, Ramon Trias Fargas, 25-27, 08005 Barcelona, Spain. E-mail: \href{mailto:gonzalo.garciadecastro@upf.edu}{\nolinkurl{gonzalo.garciadecastro@upf.edu}}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{tipa}
\usepackage{booktabs}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Cognate beginnings to bilingual lexical acquisition},
  pdfauthor={Gonzalo Garcia-Castro; Daniela S. Avila-Varela; Ignacio Castillejo; Nuria Sebastian-Galles},
  pdflang={en},
  pdfkeywords={cognate, word acquisition, vocabulary, bilingualism, item
response theory, bayesian},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\doublespacing
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognate beginnings to bilingual lexical
acquistion }
\title{Cognate beginnings to bilingual lexical acquisition}
\author{
\textbf{Gonzalo Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center
for Brain and Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}\\\\\\
\textbf{Daniela S. Avila-Varela}~\orcidlink{0000-0002-3518-8117}\\Center
for Brain and Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:avila.varela.daniela@gmail.com}{avila.varela.daniela@gmail.com}\\\\\\
\textbf{Ignacio
Castillejo}~\orcidlink{0000-0001-7445-0416}\\Departamento de
Psicología\\Universidad Autónoma de Madrid\\Madrid,
28049\\\href{mailto:jignaciocastillejo@gmail.com}{jignaciocastillejo@gmail.com}\\\\\\
\textbf{Nuria Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center
for Brain and Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Bilingual infants show equivalent developmental trajectories of lexical
acquisition compared to their monolingual peers. This is remarkable,
given the increased complexity of their linguistic input. Recent studies
suggested that bilingual vocabulary growth is boosted by the number of
cognates shared by the pair of languages being learned, and that this
facilitation effect is driven by a stronger parallel activation of
cognates during linguistic exposure, compared to non-cognates. The
mechanisms behind this facilitation are still unclear. In this study, we
capitalise on accumulator models of language acquisition to propose an
account of bilingual lexical acquisition in which parallel activation
increases the rate at which children accumulate learning instances for
words in both languages, even in fully monolingual situations. Under
this hypothesis, we predicted a stronger cognate facilitation for words
to which children were exposed less frequently (low-exposure words), as
they are co-activated by their translation more often than high-exposure
words do. We developed an extensive online vocabulary checklist, the
Barcelona Vocabulary questionnaire (BVQ) to collect vocabulary data from
366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We then
used Bayesian explanatory item response theory to model the acquisition
trajectories of 302 Catalan/Spanish translation equivalents. We found an
interaction between exposure and cognateness, which pointed to
cognateness facilitating the aquisition of low-exposure words, but not
of mean exposure or high-exposure words. Overall, our findings suggest
that cognateness plays a key role in bilingual lexical acquisition, and
provides evidence for a frequency-mediated facilitation effect driven by
parallel activation.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep word
acquisition \sep vocabulary \sep bilingualism \sep item response
theory \sep 
bayesian

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, frame hidden, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

\hypertarget{introductionauthor-note}{%
\section[Introduction]{\texorpdfstring{Introduction\footnote{The authors
  declare no conflicts of interest with regard to the funding source of
  this study. This research was supported by grants from the Spanish
  Ministerio de Ciencia, Innovación y Universidades
  (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government
  {[}ICREA (Catalan Institution for Research and Advanced Studies)
  Academia 2019 award{]}. Gonzalo Garcia-Castro was supported by a
  fellowship of the Spanish Ministerio de Ciencia, Innovación y
  Universidades (FPI 2019). We are grateful to Chiara Santolin, Ege E.
  Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest
  of the Speech Acquisition and Perception research group for their
  helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina
  Cuadrado for their technical support, and Cristina Dominguez and Katia
  Pistrin for their efforts in recruiting infants. We also thank all
  families and infants who participated in the experiments.}}{Introduction}}\label{introductionauthor-note}}

The foundations of word learning are in place early in age. At six
months, infants start directing their gaze to some objects when hearing
their labels (\protect\hyperlink{ref-bergelson2012months}{Bergelson \&
Swingley, 2012}, \protect\hyperlink{ref-bergelson2015early}{2015};
\protect\hyperlink{ref-tincoff1999beginnings}{Tincoff \& Jusczyk,
1999}), and shortly after caregivers start reporting some words as
acquired by their infant in vocabulary checklists (e.g.,
\protect\hyperlink{ref-fenson2007macarthurbates}{Fenson et al., 2007};
\protect\hyperlink{ref-samuelson2021precision}{Samuelson, 2021}). Most
research on early word acquisition relies extensively on data from
monolingual children, and is oblivious to the fact that a substantial
proportion of the world population acquires more than one language from
early ages (\protect\hyperlink{ref-grosjean2021extent}{Grosjean, 2021}).
Previous work on bilingual vocabulary acquisition pointed to bilingual
toddlers knowing, on average, less words in each of their languages than
their monolinguals peers, and to both groups knowing a similar number of
words---if not more words---when the bilinguals' two languages are
pooled together. Hoff et al.
(\protect\hyperlink{ref-hoff2012dual}{2012}) found that English-Spanish
bilingual toddlers in South Florida (United States) knew less words in
English than monolinguals did, but both groups knew a similar total
amount of words when both English and Spanish vocabularies were counted
together. Other studies have provided converging evidence that
bilinguals know a similar or even larger number of words than
monolinguals when the two languages are aggregated
(\protect\hyperlink{ref-gonzalez-barrero2020bilingual}{Gonzalez-Barrero
et al., 2020}; \protect\hyperlink{ref-oller2002language}{Oller \&
Eilers, 2002}; \protect\hyperlink{ref-patterson2004comparing}{Patterson,
2004}; \protect\hyperlink{ref-patterson2004bilingual}{Patterson \&
Pearson, 2004}; \protect\hyperlink{ref-pearson1993lexical}{Pearson et
al., 1993}; \protect\hyperlink{ref-pearson1994patterns}{Pearson \&
Fernández, 1994}; \protect\hyperlink{ref-petitto2001bilingual}{Petitto
et al., 2001}; \protect\hyperlink{ref-smithson2014bilingualism}{Smithson
et al., 2014}). A more detailed analysis of the words in bilinguals'
lexicon show some interesting patterns.

One important observation of studies on bilinguals' early vocabulary
acquisition is that cognate words are easier to acquire than non-cognate
words. Cognate words are translation equivalents that are phonologically
similar (or share some type of form-similarity). For instance, the
Spanish translation equivalent of \emph{cat} is \emph{gato}, a cognate
word; the translation equivalent of \emph{dog} is \emph{perro}, a
non-cognate word. The differences in the percentages of cognate and
non-cognate words between two languages is related to historic reasons:
languages typologically close (like Dutch and English or Italian and
Spanish) share more cognates than languages typologically distant (like
English and Chinese, or Urdu and Spanish). The conclusion that cognate
words are easier to learn is based on two types of evidence: studies
investigating vocabulary sizes in children learning language pairs with
different percentages of cognates (that is, differing in their
typological distance) and studies comparing the number of cognate and
non-cognate words children know in a specific language pair.

Floccia et al. (\protect\hyperlink{ref-floccia2018introduction}{2018})
published an impressive study comparing vocabularies of children
learning several language pairs differing in their percentage of
cognates. The authors collected vocabulary data on word comprehension
and production from 372 24-month-old bilingual toddlers living in the
United Kingdom who were learning English and an additional language. The
additional language was one of 13 typologically diverse languages:
Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu,
Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The
authors calculated the average phonological overlap between the words in
each of these additional languages and their translation equivalents in
English, which was taken as a proxy of the degree of cognateness between
each pair of languages. Floccia and co-workers reported an increase in
vocabulary size in the additional language (i.e., not English)
associated with an increase in the average phonological similarity
between the translation equivalents of each language pair. For example,
English-Dutch bilinguals (languages with a high phonological overlap),
were able to produce more Dutch words than English-Mandarin bilinguals
(languages with a low phonological overlap) were able to produce in
Mandarin. Blom et al.
(\protect\hyperlink{ref-blom2020crosslanguage}{2020}), Bosma et al.
(\protect\hyperlink{ref-bosma2019longitudinal}{2019}), and Gampe et al.
(\protect\hyperlink{ref-gampe2021does}{2021}) reported similar results,
providing converging evidence of a facilitatory effect of language
distance on vocabulary size.

A second set of studies concluded that cognates are overrepresented in
bilinguals' early lexicon. Bosch \& Ramon-Casas
(\protect\hyperlink{ref-bosch2014first}{2014}) collected parental
reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged
18 months and found that cognates represented a larger proportion of
participant's vocabulary than non-cognates. Schelletter
(\protect\hyperlink{ref-schelletter2002effect}{2002}) provided
converging evidence from a longitudinal single-case study, in which an
English-German bilingual child produced cognates earlier than
non-cognates, on average. Mitchell et al.
(\protect\hyperlink{ref-mitchell2022cognates}{2022}) addressed this
issue in a larger sample longitudinal study. These authors collected
expressive vocabulary data of 47 16- to 30-month-old French-English
bilinguals living in Canada, in both languages. They created two lists
of translation equivalents: one made of 131 cognates, and one made of
406 non-cognates. The proportion of words that children were reportedly
able to produce was higher in the cognate lists than in the non-cognate
list across ages, even when both lists were matched by semantic category
(furniture, animals, food were similarity represented in both lists) and
age-of-acquisition norms (an index of word difficulty). Taken together,
the results of these two lines of research provide a strong support to
the hypothesis that phonological similarity (as reflected in
cognateness) plays a facilitation role in bilingual word acquisition.

Parallel activation of bilinguals' lexicons has been proposed as the
underlying mechanism for such facilitatory effect (e.g.,
\protect\hyperlink{ref-floccia2018introduction}{Floccia et al., 2018};
\protect\hyperlink{ref-mitchell2022cognates}{Mitchell et al., 2022}).
The parallel activation hypothesis stems from the language non-selective
account of lexical access, which suggests that bilinguals activate both
languages simultaneously during language processing, even in fully
monolingual contexts Evidence with adult bilinguals supporting the
language-non selective account of lexical access has been reported for
language comprehension and production, across the auditory and visual
(reading and signing) modalities
(\protect\hyperlink{ref-gimeno-martinez2021crosslinguistic}{Gimeno-Martínez
et al., 2021}; \protect\hyperlink{ref-vanhell2008sentence}{Hell \&
Groot, 2008}; \protect\hyperlink{ref-hoshino2008cognate}{Hoshino \&
Kroll, 2008}; \protect\hyperlink{ref-morford2011when}{Morford et al.,
2011}; \protect\hyperlink{ref-schwartz2007reading}{Schwartz et al.,
2007}; \protect\hyperlink{ref-spivey1999cross}{Spivey \& Marian, 1999};
see \protect\hyperlink{ref-kroll2017bilingual}{Kroll \& Ma, 2017} for
review). One of the clearest pieces of evidence of parallel activation
was provided by Costa et al.
(\protect\hyperlink{ref-costa2000cognate}{2000}). In this study, Spanish
monolinguals and Catalan-Spanish bilingual adults were asked to name
pictures of common objects in Spanish. In half of the trials, the object
labels were cognates in Spanish and Catalan (\emph{árbol}-\emph{arbre},
translations of \emph{tree}), whereas in the other half of the trial
labels were non-cognates (\emph{mesa}-\emph{taula}, translations of
\emph{table}), obviously, such distinction was only relevant for
bilinguals. Bilinguals named cognate pictures faster than non-cognate
pictures, even after adjusting for the lexical frequency of the items.
In contrast, Spanish monolinguals, who were unfamiliar with the Catalan
translations of the Spanish words they uttered, showed equivalent naming
times for the two types of stimuli. The authors interpreted the
difference between cognates and non-cognates in bilinguals as reflecting
the additional phonological activation that cognate words would receive
from their translation equivalents (due to language non-selective
activation of bilinguals' lexicons). These results showed that
bilinguals' Catalan phonology was activated during the production of
Spanish words, facilitating the naming of cognate pictures. Evidence of
parallel activation has been reported in bilingual toddlers and children
(\protect\hyperlink{ref-bosma2020cognate}{Bosma \& Nota, 2020};
\protect\hyperlink{ref-poarch2012crosslanguage}{Poarch \& Hell, 2012};
\protect\hyperlink{ref-poulin-dubois2013lexical}{Poulin-Dubois et al.,
2013}; \protect\hyperlink{ref-schroter2016orthographic}{Schröter \&
Schroeder, 2016}; \protect\hyperlink{ref-vonholzen2019impact}{Von Holzen
et al., 2019}; \protect\hyperlink{ref-vonholzen2012language}{Von Holzen
\& Mani, 2012}).

Although there is a consensus on the role of parallel activation in
bilinguals' lexical processing and acquisition, previous studies do not
address its influence on the learning trajectories of words. Results are
aggregated across words and provide no information about the specific
dynamics of how parallel activation influences word learning. This is
the goal of the present research.

We propose an account in which a learning instance for a word may also
represent a learning instance for its translation equivalent, to the
extent that such translation equivalent is co-activated\footnote{We use
  the term \emph{learning instance} in the fashion of accumulator models
  of language acquisition; as an exposure to a word-form that
  constitutes an opportunity for the child to accumulate information
  about the word. We do not make strong assumptions about whether a
  learning instance is a discrete or a continuous unit of accumulation
  of information. We rather consider that a learning instance of a word
  is so to the degree of the strength of activation of its phonological
  representation. This activation may be the result of the infant being
  exposed to the word-form, or the result of activation spreading
  through phonological or semantic links across lexicl representation,
  as in the case of parallel activation.}. The strength of this
co-activation is proportional to the phonological similarity between the
two translation equivalents; given that cognates share higher
phonological similarity than non-cognates, the former should be
co-activated more frequently than the latter. This should lead to a
faster accumulation of learning instances for cognates, compared to
non-cognates. Parallel activation would allow bilingual children to
accumulate learning instances for words in both languages even during
fully monolingual situations, but the impact of this mechanism would be
asymmetric across languages: words from the lower-exposure language
would receive additional activation through parallel activation more
often than words from the higher-exposure language. Therefore, the
acquisition of words from the lower-exposure language would benefit more
strongly from their cognate status than words from the higher-exposure
language.

Consider the example of the Catalan-Spanish cognate translation
equivalent /\textipa{"gat}-\textipa{"ga.to}/ {[}cat{]}, that are
phonologically very similar. When the child is exposed to
/\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in
parallel. Therefore, this exposure will count as a learning instance for
both co-activated forms. However, for the case of the non-cognate
translation equivalent /\textipa{"gos}/, the low phonological similarity
will result in a weak activation of /\textipa{"pe.ro}/ resulting in such
exposure counting as a learning instance for /\textipa{"gos}/ (which the
child was exposed to), but not for /\textipa{"pe.ro}/. While
/\textipa{"gat}-\textipa{"ga.to}/ will benefit from parallel activation,
/\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives
linguistic input from one of the languages more often than from the
other, this effect might affect each form of the cognate translation
equivalent differently. For instance, if the child receives a larger
amount of Catalan input than Spanish input, they will encounter the
Catalan form /\textipa{"gat}/ more frequently than the Spanish form
/\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will
activate /\textipa{"ga.to}/ more often than vice versa. Ultimately,
/\textipa{"ga.to}/ will benefit more strongly from its cognate status
than /\textipa{"gat}/, as it receives additional learning instances from
its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production
and comprehension from a large sample of bilingual Catalan-Spanish
children using an online vocabulary checklist designed for the present
study. We adopted a Bayesian explanatory item response theory approach
(IRT, see \protect\hyperlink{ref-kachergis2022standard}{Kachergis et
al., 2022}, for a similar approach) to model the probability of
acquisition of 604 Catalan and Spanish nouns included in the vocabulary
checklist. Words were considered as acquired if caregivers reported such
word to be understood (comprehension) or understood and said
(production) by their child. We estimated the impact of several
predictors of interest on the probability of acquisition, including
participants' age and rate of exposure to the word-form, and the cognate
status of the word-form. As described in the methods section, rate of
exposure was a composite measure taking into account participant'
language exposure and word's lexical frequency. We predicted an
interaction between cognate status and word exposure rate in which the
probability of comprehension is higher for low-exposure cognate words,
but not for high-exposure words cognate words.

\hypertarget{sec-methods}{%
\section{Methods}\label{sec-methods}}

All materials, data, and reproducible code can be found at the OSF
(\url{https://osf.io/hy984/}) and GitHub
(\url{https://github.com/gongcastro/cognate-beginnings}) repositories.
This study was conducted according to guidelines laid down in the
Declaration of Helsinki, and was approved by the Drug Research Ethical
Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I.

\hypertarget{sec-questionnaire}{%
\subsection{Questionnaire}\label{sec-questionnaire}}

To collect vocabulary data from participants, we created an \emph{ad
hoc} questionnaire: the Barcelona Vocabulary Questionnaire (BVQ)
(\protect\hyperlink{ref-garcia-castro2023bvq}{Garcia-Castro et al.,
2023}). This questionnaire was inspired by the MacArthur-Bates
Communicative Development Inventory
(\protect\hyperlink{ref-fenson2007macarthurbates}{Fenson et al., 2007})
and its adaptations to other languages, and was implemented on-line
using the formr platform (\protect\hyperlink{ref-arslan2020formr}{Arslan
et al., 2020}). This questionnaire is structured in three blocks: (1) a
language exposure questionnaire, (2) a demographic survey, and (3) two
vocabulary checklists. Vocabulary checklists followed a similar
structure as the Oxford Communicative Developmental Inventory (OCDI)
(\protect\hyperlink{ref-hamilton2000infant}{Hamilton et al., 2000}) and
consisted in two lists of words: one in Catalan and one in Spanish. Both
lists included items from a diverse sample of 26 semantic/functional
categories. The Catalan checklist contained 793 items and the Spanish
checklist contained 797. Items in one language were translation
equivalents of the items in the other language (e.g., the same
participant responded to both \emph{gos} and \emph{perro}, Catalan and
Spanish for \emph{dog}), roughly following a one-to-one
mapping\footnote{Although for some of the included words in Catalan did
  not have a clear translation or had more than one possible translation
  in Spanish, and vice versa, therefore the unequal number of words
  included in the Catalan and Spanish lists} (see Table~\ref{tbl-items}
for a summary of the questionnaire items).

\hypertarget{tbl-items}{}
\begin{longtable}{l|rrrrc}
\caption{\label{tbl-items}Summary of the items included in the final analyses. }\tabularnewline
